---
title: "Cell Type Classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cell Type Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

scMMR's core functionality is a multi-task deep neural network that jointly
learns **cell type classification** and **UMAP embedding regression** from a
shared ResNet backbone. This article covers the full workflow from training to
interpretability.

## Architecture

```
Input (binarized expression: X > 0)
  |
InputBlock: BatchNorm -> Dropout(0.25) -> Linear(n_genes -> 512) -> SELU
  |
ResNetBlock x4: Dropout(0.1) -> Linear(512 -> 512) -> SELU + skip connection
  |
SharedEmbedding (512-dim)
  |--- ClassificationHead -> Linear(512->256) -> SELU -> Linear(256->K)
  |--- RegressionHead -> Linear(512->256) -> SELU -> Linear(256->D)
```

Key features:
- **GradNorm** for multi-task loss balancing
- **Label smoothing** regularization
- **OOD detection** via confidence threshold
- **Integrated Gradients** for gene importance

## Training

### Prepare Reference Data

The training function accepts either an h5ad file path or a Seurat object:

```{r}
library(scMMR)
use_scMMR_python(condaenv = "/path/to/conda/env")

# From Seurat object
train_result <- DNN_train(
  input       = seurat_ref,
  label_col   = "cell_type",
  umap_cols   = c("umap_1", "umap_2"),
  n_top_genes = 3000,
  batch_key   = "batch",
  num_epochs  = 50,
  device      = "cuda"
)
```

### Key Training Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `n_top_genes` | 3000 | Number of highly variable genes |
| `batch_key` | NULL | Batch column for HVG selection |
| `num_epochs` | 50 | Training epochs |
| `learning_rate` | 1e-3 | Initial learning rate |
| `hidden_size` | 512 | Hidden layer width |
| `ood_threshold` | 0.5 | OOD detection confidence cutoff |
| `device` | "cpu" | "cpu" or "cuda" |

### Training Output

```{r}
# Trained model artifacts
train_result$model_dir    # Path to saved model.pt + var_genes.json
train_result$history      # Training curves (loss, accuracy, RMSE per epoch)
train_result$best_val_acc # Best validation accuracy
```

## Prediction

```{r}
result <- DNN_predict(
  input         = seurat_query,
  model_dir     = train_result$model_dir,
  explain       = TRUE,
  top_k_global  = 30,
  top_k_class   = 20,
  pathway_gmt   = system.file("extdata/gmt/h.all.v2022.1.Hs.symbols.gmt",
                               package = "scMMR"),
  return_embedding = TRUE,
  device        = "cpu"
)
```

### Prediction Output

| Component | Type | Description |
|-----------|------|-------------|
| `result$predictions` | data.frame | cell_type_pred, confidence, is_ood |
| `result$shared_embedding` | matrix | 512-dim embedding (n_cells x 512) |
| `result$imp_global` | data.frame | Top genes by global importance |
| `result$imp_per_class` | list | Per-class gene importance |
| `result$pathway_scores` | data.frame | Per-cell pathway activity scores |

## Merge Predictions into Seurat

```{r}
library(Seurat)

# Add predictions
seu <- AddMetaData(seu, result$predictions)

# Add embedding as a new reduction
emb <- result$shared_embedding
rownames(emb) <- colnames(seu)
seu[["dnn"]] <- CreateDimReducObject(emb, key = "DNN_")
```

## Visualizations

### UMAP Projection (PlotMAP)

Overlay query predictions on a reference UMAP:

```{r}
ref_umap <- qs::qread(system.file("extdata", "ref_umap.qs", package = "scMMR"))

PlotMAP(
  ref       = ref_umap,
  query_meta = seu@meta.data,
  color_by  = "cell_type_pred",
  ref_emb   = c("umap_1", "umap_2"),
  query_emb = c("umap_pred_1", "umap_pred_2")
)
```

### Cell Type Composition (PlotAlluvia / PlotSankey)

```{r}
# Stacked area + bar chart
PlotAlluvia(seu, by = "group", fill = "cell_type_pred")

# Sankey flow diagram
PlotSankey(seu, by = "group", fill = "cell_type_pred")
```

### Gene Importance (PlotImportance)

```{r}
# Global gene importance (top 30 genes)
PlotImportance(result$imp_global, top_k = 30, display = "lollipop")

# Per-group importance
PlotImportance(result$imp_per_class, top_k = 15, display = "bar")
```

### Annotation Summary (PlotAnnotation)

```{r}
PlotAnnotation(
  seu,
  which = c("cell_type_pred", "group", "confidence"),
  palette = "Set2"
)
```

## OOD Detection

Out-of-distribution cells are flagged when prediction confidence falls below
the threshold:

```{r}
table(seu$is_ood)
# FALSE  TRUE
# 4800    200

# Filter high-confidence predictions
seu_hc <- subset(seu, is_ood == FALSE)
```

## Tips

- Use `device = "cuda"` for GPU acceleration (10-50x faster on large datasets)
- Set `batch_key` during training if your reference has batch effects
- For cross-dataset prediction, the model automatically intersects gene lists
- Increase `n_top_genes` (e.g., 5000) for higher resolution at the cost of speed
