---
title: "Bulk RNA-seq Deconvolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bulk RNA-seq Deconvolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

scMMR provides DNN-based bulk RNA-seq deconvolution that estimates cell type
proportions from bulk expression profiles. The model is trained on
pseudo-bulk samples generated from scRNA-seq reference data using Dirichlet
sampling.

## Architecture

```
Input (log1p-CPM, continuous)
  |
InputBlock -> ResNetBlock x4 -> SharedEmbedding (512-dim)
  |
TaskHead -> Linear(512->256) -> SELU -> Linear(256->K) -> Softmax
  |
Proportions (sum = 1)
```

Key difference from classification DNN:
- **Input**: Continuous log-CPM (not binarized)
- **Output**: Cell type proportion vector (softmax-normalized)
- **Training data**: Dirichlet-sampled pseudo-bulk mixtures

## Step 1: Train on scRNA-seq Reference

```{r}
library(scMMR)
use_scMMR_python(condaenv = "/path/to/conda/env")

# Using Seurat object as reference
train_result <- DNN_deconv_train(
  input              = seurat_ref,
  label_col          = "cell_type",
  n_pseudobulk       = 5000,
  n_cells_per_sample = 500,
  n_top_genes        = 3000,
  loss_function      = "mse",
  num_epochs         = 100,
  device             = "cuda"
)
```

### Training Parameters

| Parameter | Default | Description |
|-----------|---------|-------------|
| `n_pseudobulk` | 5000 | Number of pseudo-bulk samples to generate |
| `n_cells_per_sample` | 500 | Cells per pseudo-bulk sample |
| `n_top_genes` | 3000 | Highly variable genes for training |
| `loss_function` | "mse" | Loss: "mse" or "kl" |
| `recon_weight` | 0.0 | Weight for reconstruction auxiliary loss |
| `num_epochs` | 100 | Training epochs |
| `device` | "cpu" | "cpu" or "cuda" |

### Training Output

```{r}
train_result$model_dir   # Path to model.pt + var_genes.json + cell_types.json
train_result$history      # Training/validation loss curves
train_result$cell_types   # Character vector of cell types
```

## Step 2: Predict Cell Type Proportions

### Standard Prediction

```{r}
# From bulk expression matrix (genes x samples)
proportions <- DNN_deconv_predict(
  input     = bulk_matrix,
  model_dir = train_result$model_dir,
  device    = "cpu"
)

head(proportions)
#>         acinar   alpha    beta   delta  ductal endothelial
#> Sample1  0.12   0.35    0.28   0.08    0.10       0.03
#> Sample2  0.08   0.42    0.20   0.05    0.15       0.04
```

### Adaptive Deconvolution

Adaptive mode iteratively refines predictions by extracting cell-type-specific
gene expression profiles (GEP):

```{r}
result <- DNN_deconv_predict(
  input         = bulk_matrix,
  model_dir     = train_result$model_dir,
  adaptive      = TRUE,
  adaptive_mode = "overall",
  adaptive_max_iter = 100,
  device        = "cpu"
)

result$proportions  # Refined proportions
result$sigmatrix    # Estimated cell-type-specific GEP matrix
```

### Supported Input Formats

| Format | Example |
|--------|---------|
| Matrix (genes x samples) | `DNN_deconv_predict(input = mat)` |
| CSV file path | `DNN_deconv_predict(input = "bulk.csv")` |
| h5ad file path | `DNN_deconv_predict(input = "bulk.h5ad")` |
| Seurat object | `DNN_deconv_predict(input = seurat_bulk)` |

## Step 3: Evaluate Results

### Correlation with Ground Truth

```{r}
# If true proportions are known (e.g., from pseudo-bulk testing)
cor_results <- sapply(colnames(proportions), function(ct) {
  cor(proportions[[ct]], true_proportions[[ct]], method = "pearson")
})

# Per-cell-type RMSE
rmse_results <- sapply(colnames(proportions), function(ct) {
  sqrt(mean((proportions[[ct]] - true_proportions[[ct]])^2))
})
```

### Visualization

```{r}
library(ggplot2)

# Scatter: predicted vs true proportions
ggplot(merged_df, aes(x = true, y = predicted)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ cell_type, scales = "free") +
  labs(x = "True Proportion", y = "Predicted Proportion")

# Stacked bar chart
ggplot(prop_long, aes(x = sample, y = proportion, fill = cell_type)) +
  geom_col() +
  scale_fill_manual(values = palette_colors(cell_types)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Tips

- Use raw counts or CPM as input (not log-transformed) --
  the model applies `log1p(CPM)` internally
- Increase `n_pseudobulk` for better training (5000-10000 recommended)
- Use `loss_function = "kl"` if proportions are very skewed
- Adaptive mode is slower but more accurate for heterogeneous samples
