---
title: "Getting Started with scMMR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with scMMR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

**scMMR** (Single-Cell Multi-task Model in R) is a comprehensive toolkit for
single-cell RNA-seq analysis powered by PyTorch multi-task deep neural networks.
It provides:
- Joint cell type classification and embedding regression
- Bulk RNA-seq deconvolution with cell type proportion estimation
- Gene importance attribution via Integrated Gradients
- Pathway enrichment analysis (GSEA, SCPA)
- Perturbation ranking with distribution-distance metrics
- Pseudotime trajectory analysis and cellular potency prediction
- A rich visualization suite (18+ publication-ready plot functions)

## Installation

### Step 1: Install scMMR from GitHub

```{r}
if (!requireNamespace("devtools", quietly = TRUE))
  install.packages("devtools")
devtools::install_github("HUI950319/scMMR")
```

### Step 2: Set Up Python Environment

scMMR uses PyTorch via `reticulate`. You can either create a fresh conda
environment or point to an existing one.

**Option A: Automatic installation** (creates a new conda environment):

```{r}
library(scMMR)
install_scMMR_python(envname = "scMMR-env", method = "conda")
```

This installs: `torch`, `scanpy`, `anndata`, `numpy`, `pandas`, `scipy`, `h5py`.

**Option B: Use an existing conda environment**:

```{r}
library(scMMR)
use_scMMR_python(condaenv = "/path/to/your/conda/env")
```

### Step 3: Verify Setup

```{r}
library(scMMR)
use_scMMR_python(condaenv = "/path/to/your/conda/env")

# Check that Python modules are accessible
reticulate::py_module_available("torch")
reticulate::py_module_available("scanpy")
```

## Quick Start: Predict Cell Types

Here is a minimal example using the bundled test data and pre-trained model:

```{r}
library(scMMR)
library(Seurat)

# 1. Activate Python environment
use_scMMR_python(condaenv = "/path/to/your/conda/env")

# 2. Load test data and model
seu <- qs::qread(system.file("extdata", "toy_test.qs", package = "scMMR"))
model_path <- system.file("extdata", "model.pt", package = "scMMR")

# 3. Run prediction with gene importance
result <- DNN_predict(
  input     = seu,
  model_dir = dirname(model_path),
  explain   = TRUE,
  top_k_global = 20,
  device    = "cpu"
)

# 4. Merge predictions into Seurat object
seu <- AddMetaData(seu, result$predictions)

# 5. Visualize cell type composition
PlotAlluvia(seu, by = "group", fill = "cell_type_pred")
```

## What's Next?

- [Cell Type Classification](cell-type-classification.html) -- Train and predict
  with multi-task DNN
- [Bulk Deconvolution](deconvolution.html) -- Estimate cell type proportions
  from bulk RNA-seq
- [Pathway Analysis](pathway-analysis.html) -- GSEA and SCPA enrichment
- [Perturbation Ranking](perturbation-ranking.html) -- Rank cell types by
  treatment response
- [Visualization Gallery](visualization-gallery.html) -- All 18+ plot functions
