---
title: "Gene Set Module Scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene Set Module Scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

`ComputeModuleScore()` computes per-cell gene set activity scores using three
methods: **AUCell**, **Seurat AddModuleScore**, and **UCell**. It is an S3
generic that works with both raw matrices and Seurat objects.

## Supported Input Formats

Gene sets can be provided in three formats:

### 1. Named List

```{r}
gene_sets <- list(
  "Hallmark_TNFA"   = c("JUNB", "CXCL2", "ATF3", "NFKBIA", "TNFAIP3"),
  "Hallmark_HYPOXIA" = c("VEGFA", "SLC2A1", "ALDOA", "ENO1", "LDHA")
)
```

### 2. Data Frame (term-gene pairs)

```{r}
gene_sets_df <- data.frame(
  term = c(rep("TNFA_SIGNALING", 5), rep("HYPOXIA", 5)),
  gene = c("JUNB", "CXCL2", "ATF3", "NFKBIA", "TNFAIP3",
           "VEGFA", "SLC2A1", "ALDOA", "ENO1", "LDHA")
)
```

### 3. GMT File Path

```{r}
gmt_path <- system.file("extdata/gmt/h.all.v2022.1.Hs.symbols.gmt",
                         package = "scMMR")
```

## Scoring Methods

### AUCell (Default)

Area Under the recovery Curve -- ranks genes by expression and computes
the AUC of gene set members in the top-ranked genes:

```{r}
scores <- ComputeModuleScore(
  object    = expression_matrix,  # genes x cells
  gene.sets = gene_sets,
  method    = "AUCell"
)
# Returns: pathways x cells matrix
```

### Seurat AddModuleScore

Computes the average expression of gene set genes minus the average of
control gene sets:

```{r}
scores <- ComputeModuleScore(
  object    = expression_matrix,
  gene.sets = gene_sets,
  method    = "Seurat",
  nbin      = 24,
  ctrl      = 100
)
```

### UCell

Mann-Whitney U statistic-based scoring, robust to library size differences:

```{r}
scores <- ComputeModuleScore(
  object    = expression_matrix,
  gene.sets = gene_sets,
  method    = "UCell"
)
```

## Using with Seurat Objects

The Seurat method wraps the default and stores results directly:

```{r}
library(Seurat)

# Store as a new assay (recommended for multi-pathway scoring)
seu <- ComputeModuleScore(
  object    = seu,
  gene.sets = gmt_path,
  method    = "AUCell",
  store     = "assay",
  assay.name = "Hallmark"
)

# Access scores
seu[["Hallmark"]]@data[1:3, 1:5]

# Store as metadata columns (simpler for a few pathways)
seu <- ComputeModuleScore(
  object    = seu,
  gene.sets = gene_sets,
  method    = "Seurat",
  store     = "metadata",
  prefix    = "Score_"
)

head(seu@meta.data[, grep("Score_", colnames(seu@meta.data))])
```

## Working with GMT Files

### Read and Parse GMT

```{r}
# Read GMT file
gs <- read_gmt(system.file("extdata/gmt/h.all.v2022.1.Hs.symbols.gmt",
                            package = "scMMR"))
length(gs)  # 50 Hallmark gene sets
names(gs)[1:5]

# Flexible parsing (auto-detects format)
gs2 <- parse_gene_sets(gmt_path)                  # GMT file
gs3 <- parse_gene_sets(gene_sets)                  # Named list (pass-through)
gs4 <- parse_gene_sets(gene_sets_df)               # Data frame
```

### Filter by Gene Set Size

```{r}
# Keep only gene sets with 15-500 genes
gs_filtered <- gs[sapply(gs, length) >= 15 & sapply(gs, length) <= 500]
```

## Bundled GMT Databases

scMMR ships with 14 GMT files covering major pathway databases:

| Database | Species | File | Gene Sets |
|----------|---------|------|-----------|
| MSigDB Hallmark | Human | `h.all.v2022.1.Hs.symbols.gmt` | 50 |
| KEGG Legacy | Human | `c2.cp.kegg.v2022.1.Hs.symbols.gmt` | 186 |
| Reactome | Human | `reactome.gmt` | 1,000+ |
| GO Biological Process | Human | `GO_bp.gmt` | 7,000+ |
| CollecTRI (regulons) | Human | `collectri.human.gmt` | 700+ |
| PROGENy (signaling) | Human | `progeny.human.top100.gmt` | 14 |
| Proliferation | Human | `proliferation.combined.gmt` | Custom |
| Reactome | Mouse | `m_reactome.gmt` | 1,000+ |
| GO Biological Process | Mouse | `m_GO_bp.gmt` | 7,000+ |

## Comparison of Methods

| Feature | AUCell | Seurat | UCell |
|---------|--------|--------|-------|
| Score range | [0, 1] | Unbounded | [0, 1] |
| Speed | Moderate | Fast | Moderate |
| Library size robust | Yes | Partial | Yes |
| Best for | Regulatory activity | Quick exploration | Small gene sets |

## Tips

- Use `method = "AUCell"` for gene regulatory network scoring
- Use `method = "Seurat"` for quick exploratory analysis
- Use `method = "UCell"` when comparing across batches with different depths
- Filter gene sets to 15-500 genes for meaningful scores
- Use `store = "assay"` when scoring many pathways to keep metadata clean
