---
title: "Perturbation Ranking & Differential Abundance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perturbation Ranking & Differential Abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

scMMR provides two complementary approaches for ranking cell types by their
response to perturbation (e.g., treatment vs control):

1. **RankPerturbation** -- Measures distribution distances in embedding space
   (how much each cell type "moves")
2. **RankPercent** -- Tests differential abundance using neighborhood-based
   permutation (whether cell type proportions change)

## Cell Composition Visualization

Before diving into statistical tests, visualize the overall composition:

### O/E Ratio Heatmap (PlotRoe)

The Observed/Expected (O/E) ratio reveals group-specific enrichment of each
cell type. Values > 1 indicate enrichment, < 1 indicates depletion.

```{r}
# Basic O/E heatmap
PlotRoe(seu, by = "group", fill = "cell_type")

# With significance stars (chi-squared test)
PlotRoe(seu, by = "group", fill = "cell_type",
        show.pvalue = TRUE, p.method = "chisq")

# With grade symbols (+++ / ++ / + / +/- / -)
PlotRoe(seu, by = "group", fill = "cell_type",
        display = "symbol")

# Both value and symbol
PlotRoe(seu, by = "group", fill = "cell_type",
        display = "both", show.pvalue = TRUE)

# Return data for custom analysis
roe_data <- PlotRoe(seu, by = "group", fill = "cell_type",
                     show.pvalue = TRUE, return.data = TRUE)
roe_data$oe_matrix    # O/E ratio matrix
roe_data$pval_matrix  # P-value matrix
```

### Alluvial / Sankey Diagrams

```{r}
# Stacked area chart showing composition across groups
PlotAlluvia(seu, by = "group", fill = "cell_type")

# Sankey flow diagram
PlotSankey(seu, by = "group", fill = "cell_type",
           show.pct = TRUE, show.label = TRUE)
```

## RankPerturbation: Distribution Distance

RankPerturbation measures how much each cell type's embedding distribution
shifts between conditions using five distance metrics:

| Metric | Description | Best for |
|--------|-------------|----------|
| `wasserstein` | Earth mover's distance (1D projections) | General use |
| `sliced_wasserstein` | Multi-dimensional Wasserstein | High-dim embeddings |
| `mmd` | Maximum Mean Discrepancy (RBF kernel) | Non-linear shifts |
| `energy` | Energy distance (non-parametric) | Heavy-tailed distributions |
| `auc` | Classifier-based AUC (LDA) | Separability measure |

### Basic Usage

```{r}
# Run perturbation ranking
perturb <- RankPerturbation(
  embedding    = result$shared_embedding,  # n_cells x 512
  cell_meta    = seu@meta.data,
  celltype_col = "cell_type",
  conditions   = list(
    treatment = "group == 'Treatment'",
    control   = "group == 'Control'"
  ),
  method       = "mean",
  alpha        = 0.1,
  n_perm       = 1000,
  seed         = 42
)
```

### Output

```{r}
head(perturb$results)
#>   cell_type   score   p_value  p_adjust  rank
#> 1 T_cell      0.85    0.001    0.008     1
#> 2 Macrophage  0.72    0.003    0.015     2
#> 3 Fibroblast  0.45    0.045    0.120     3
```

### Visualize Rankings

```{r}
# Lollipop chart (default)
PlotPerturbation(perturb, top_k = 10, display = "lollipop")

# Bar chart with significance stars
PlotPerturbation(perturb, top_k = 15, display = "bar",
                  sig_threshold = 0.05)
```

## RankPercent: Differential Abundance

RankPercent performs neighborhood-based differential abundance testing,
similar to miloR. It constructs KNN/SNN graphs and tests whether cell type
proportions differ between conditions.

### Basic Usage

```{r}
da <- RankPercent(
  embedding          = result$shared_embedding,
  cell_meta          = seu@meta.data,
  celltype_col       = "cell_type",
  conditions         = list(
    treatment = "group == 'Treatment'",
    control   = "group == 'Control'"
  ),
  neighborhood_method = "knn",
  k_neighborhoods     = 30,
  n_perm              = 1000,
  fdr_threshold       = 0.1
)
```

### Output

```{r}
head(da$results)
#>   cell_type   logFC    p_value  p_adjust  sig
#> 1 T_cell      1.25     0.001    0.008     TRUE
#> 2 Macrophage  -0.85    0.012    0.048     TRUE
#> 3 B_cell      0.32     0.234    0.468     FALSE
```

### Visualize Differential Abundance

```{r}
# Beeswarm plot (miloR style)
PlotPercent(da, fdr_threshold = 0.1)

# With boxplot overlay
PlotPercent(da, fdr_threshold = 0.05, show_boxplot = TRUE)
```

## Combining Both Approaches

Use both methods together for a comprehensive view:

```{r}
# 1. Which cell types change their internal state? (RankPerturbation)
perturb <- RankPerturbation(embedding, cell_meta, ...)

# 2. Which cell types change their abundance? (RankPercent)
da <- RankPercent(embedding, cell_meta, ...)

# 3. Overlay composition (PlotRoe)
PlotRoe(seu, by = "group", fill = "cell_type",
        show.pvalue = TRUE, display = "both")
```

## Tips

- RankPerturbation detects **state changes** (gene expression shifts within a
  cell type), while RankPercent detects **abundance changes** (proportion shifts)
- Use DNN embedding (512-dim) rather than PCA for more sensitive detection
- Set `n_perm >= 1000` for reliable p-values
- Use `alpha = 0.1` (default) to add noise for robust distance estimation
