---
title: "Visualization Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = FALSE,
  fig.width = 8, fig.height = 5
)
```

## Overview

scMMR includes 18+ publication-ready visualization functions. This gallery
provides a quick reference for each function with usage examples.

---

## Composition & Flow

### PlotRoe -- O/E Ratio Heatmap

Observed/Expected ratio heatmap highlighting cell type enrichment per group.

```{r}
# Numeric values
PlotRoe(seu, by = "group", fill = "cell_type")

# Grade symbols (+++ / ++ / + / +/- / -)
PlotRoe(seu, by = "group", fill = "cell_type", display = "symbol")

# With significance stars
PlotRoe(seu, by = "group", fill = "cell_type",
        show.pvalue = TRUE, p.method = "chisq")

# Both value + symbol
PlotRoe(seu, by = "group", fill = "cell_type",
        display = "both", show.pvalue = TRUE)
```

**Key parameters:** `by`, `fill`, `palette`, `display` ("value"/"symbol"/"both"),
`show.pvalue`, `p.method` ("chisq"/"fisher"), `return.data`

### PlotAlluvia -- Stacked Area + Bar Chart

Cell type composition as stacked areas and bar charts across groups.

```{r}
PlotAlluvia(seu, by = "group", fill = "cell_type")

# Customize
PlotAlluvia(seu, by = "group", fill = "cell_type",
            show.pct = TRUE, show.label = TRUE,
            bar.width = 0.8, flow.alpha = 0.5)
```

**Key parameters:** `by`, `fill`, `colors`, `palette`, `flow.alpha`, `bar.width`,
`show.label`, `show.pct`

### PlotSankey -- Sankey Flow Diagram

Alluvial-style Sankey diagram for cell transitions.

```{r}
PlotSankey(seu, by = "group", fill = "cell_type")

PlotSankey(seu, by = "group", fill = "cell_type",
           curve.type = "spline", show.pct = TRUE)
```

**Key parameters:** Same as PlotAlluvia plus `curve.type` ("linear"/"spline")

---

## Projection & Spatial

### PlotMAP -- UMAP Projection

Overlay query predictions on reference UMAP background.

```{r}
PlotMAP(
  ref       = ref_umap_df,
  query_meta = seu@meta.data,
  color_by  = "cell_type_pred",
  ref_emb   = c("umap_1", "umap_2"),
  query_emb = c("umap_pred_1", "umap_pred_2")
)
```

**Key parameters:** `ref`, `query_meta`, `ref_emb`, `query_emb`, `color_by`,
`facet_by`, `colors`, `density.color`

### PlotScatter -- General Scatter Plot

Flexible scatter plot for embeddings or feature-feature comparisons.

```{r}
PlotScatter(seu, reduction = "umap", color.by = "cell_type")

PlotScatter(seu, features = c("gene_A", "gene_B"),
            group.by = "cell_type")
```

**Key parameters:** `reduction`, `features`, `group.by`, `split.by`,
`color.by`, `colors`, `base_size`

### PlotAnnotation -- Cell Annotation Heatmap

Summary heatmap with automatic color coding for metadata columns.

```{r}
PlotAnnotation(seu, which = c("cell_type", "group", "confidence"))
```

**Key parameters:** `which`, `palette`, `row_order`, `col_order`, `base_size`

---

## Gene Importance & Ranking

### PlotImportance -- Gene Importance Chart

Lollipop or bar chart of gene importance from Integrated Gradients.

```{r}
# Global importance
PlotImportance(result$imp_global, top_k = 30, display = "lollipop")

# Per-class importance
PlotImportance(result$imp_per_class, top_k = 15, display = "bar")
```

**Key parameters:** `top_k`, `display` ("lollipop"/"bar"), `palette`, `base_size`

### PlotRankScatter -- Rank Scatter Plot

Generic scatter for visualizing feature scores (gene importance, module scores,
pathway NES).

```{r}
PlotRankScatter(scores_df, x_label = "Gene", y_label = "Importance",
                highlight = top_genes, label.top_k = 10)
```

**Key parameters:** `x_label`, `y_label`, `show_line`, `highlight`,
`label.top_k`, `colors`

---

## Perturbation & Abundance

### PlotPerturbation -- Perturbation Score Chart

Ranked chart of perturbation scores with significance markers.

```{r}
PlotPerturbation(perturb_result, top_k = 10, display = "lollipop")
PlotPerturbation(perturb_result, display = "bar", sig_threshold = 0.05)
```

**Key parameters:** `top_k`, `display` ("lollipop"/"bar"), `sig_threshold`,
`palette`, `base_size`

### PlotPercent -- Differential Abundance Beeswarm

miloR-style beeswarm plot of per-neighborhood logFC.

```{r}
PlotPercent(da_result, fdr_threshold = 0.1)
PlotPercent(da_result, show_boxplot = TRUE, point_size = 1.5)
```

**Key parameters:** `fdr_threshold`, `colors`, `show_boxplot`, `base_size`,
`point_size`

---

## Differential Expression

### PlotDE -- DE Dot Plot / Volcano

```{r}
# Dot plot
PlotDE(de_results, type = "dot", top_n = 8)

# Volcano plot
PlotDE(de_results, type = "volcano", logfc.cutoff = 1, label = TRUE)
```

**Key parameters:** `type` ("dot"/"volcano"), `top_n`, `top_by`, `logfc.cutoff`,
`pval.cutoff`, `label`

### PlotGsea -- GSEA Bubble Plot

```{r}
PlotGsea(gsea_results, top.n = 10)
PlotGsea(gsea_results, facet_by = "sign", order_by = "NES")
```

**Key parameters:** `top.n`, `order_by`, `use.padj`, `facet_by`, `colors`

---

## Pathway & Enrichment

### PlotPathwayBubble -- Pathway Enrichment Bubble

```{r}
PlotPathwayBubble(pathway_results, top.n = 10)

PlotPathwayBubble(pathway_results, top.n = 15,
                   size.by = "pvalue", facet.by.sign = TRUE)
```

**Key parameters:** `top.n`, `size.by`, `pvalue.cutoff`, `select.by`,
`order.by`, `facet.by.sign`, `colors`

---

## Correlation

### PlotCorrelation -- Correlation Scatter

Volcano-style plot: -log10(p) vs correlation coefficient.

```{r}
PlotCorrelation(cor_results, p.cutoff = 0.05, cor.cutoff = 0.3,
                label = TRUE, label.n = 10)
```

**Key parameters:** `use.padj`, `p.cutoff`, `cor.cutoff`, `label`, `label.n`

### PlotPropCorrelation -- Proportion Correlation Heatmap

```{r}
PlotPropCorrelation(seu, emb.cols = paste0("DNN_", 1:10),
                     prop.data = proportions, method = "spearman")
```

**Key parameters:** `emb.cols`, `prop.data`, `method`, `group.by`, `base_size`

---

## Trajectory & Potency

### PlotCytoTRACE2 -- Cellular Potency Plot

Violin + box + jitter plot of CytoTRACE2 scores grouped by cell type.

```{r}
PlotCytoTRACE2(seu, group.by = "cell_type",
               score = "CytoTRACE2_Score", order = TRUE)
```

**Key parameters:** `group.by`, `score`, `order`, `colors`, `base_size`

### PlotDynamicFeatures -- Feature Dynamics Along Pseudotime

```{r}
PlotDynamicFeatures(seu, features = c("gene_A", "gene_B"),
                     pseudotime_col = "pseudotime",
                     method = "loess")
```

**Key parameters:** `features`, `pseudotime_col`, `method` ("loess"/"gam"),
`group.by`, `colors`

---

## Embedding Quality

### PlotEmbeddingEval -- Embedding Evaluation Diagnostics

Five-panel diagnostic plot for DNN embedding quality.

```{r}
eval_result <- EvaluateEmbedding(embedding, seurat_obj = seu)

# All plots
PlotEmbeddingEval(eval_result, which = "all")

# Individual plots
PlotEmbeddingEval(eval_result, which = "elbow")
PlotEmbeddingEval(eval_result, which = "knn")
PlotEmbeddingEval(eval_result, which = "silhouette")
```

**Key parameters:** `which` ("all"/"elbow"/"cumvar"/"knn"/"distcor"/"silhouette"),
`base_size`

---

## Color Palettes

### palette_colors -- Generate Color Vectors

```{r}
# Automatic palette for a character vector
cols <- palette_colors(cell_types)

# Specific palette
cols <- palette_colors(n = 8, palette = "Set2")

# Continuous palette
cols <- palette_colors(n = 100, type = "continuous", palette = "RdYlBu")
```

### show_palettes -- Display Available Palettes

```{r}
# Show all discrete palettes
show_palettes(type = "discrete")

# Show specific palettes
show_palettes(palette_names = c("Set2", "Paired", "lancet"))
```

**Available palettes:** All RColorBrewer palettes + ggsci palettes
(lancet, nejm, jama, jco, d3, etc.) + custom scMMR palettes.
